R Notebook

### 2024-07-31 Scripts/authoritarian_only_latex_2.Rmd
Edits to condense and clarify

The stakes of this debate are significant. The former view is an optimistic one; incumbents avoid fraud when it might be exposed and reveal their weakness to strategic protesters\textemdash an auspicious prediction for democratization and ultimately the under-pinning of 'self-enforcing' democracy [@fearon_self-enforcing_2011]. If the latter view holds, electoral protest may mainly occur non-strategically, and would typically fail to deter election manipulation as a result. Mass protest would, in this case, be a fragile guardrail against autocratization. This paper argues that this pessimistic model best fits the overall pattern of the data, and is supported by specific, novel hypothesis tests.

This paper contributes to this debate by drawing out the implications of three strands of the literature on election manipulation and protest. In short: there are two views on the nature of electoral protest which make the same prediction. If opposition groups strategically initiate protest when they believe the incumbent to be weak, they will be most likely to do so when the election results show a low vote-share for the incumbent combined with high levels of manipulation\textemdash a combination of factors which reveals the incumbent's very low genuine popularity. However, if protest movements are more organic and non-strategic, reflecting citizen grievances, we should expect protest initiation in the same circumstances. Here the combination of low incumbent vote-share and high manipulation indicates severe citizen grievances (low natural popularity) which are galvanized by the shared grievance of a 'stolen' election. Since both theories make the same prediction about protest initiation, such tests\textemdash which have been the dominant empirical measure in the literature\textemdash do not reveal whether protest is strategic and manipulation indicates weakness or manipulation indicates sources of strength and anti-manipulation protest is non-strategic.

To unravel this question, I make use of data on the duration of post-election protests. Shorter-lived protest movements that end with the incumbent still in power indicate a government that is relatively strong compared to the protesting opposition, while longer-lived protest movements indicate a more even balance of power between the two. The reverse is also true: shorter-lived protest movements that end in opposition victory indicate a weaker government, while governments that hold on longer before being defeated are relatively stronger. Using data from 614 non-democratic elections across 97 countries, from 1989 to 2012, this paper finds that 1) protest initiation is most likely when manipulation is severe but incumbent vote-share is low, as both theories predict, and 2) that governments that engage in more manipulation are likely to put down protests more quickly than governments that produce little manipulation, especially when incumbents claim small vote-shares.

@@

Three understandings of the relationship between election manipulation and post-election protest have developed in the literature. Each view is discussed in detail below, after a brief initial summary here. A first view assumes that incumbents and opposition groups are strategic actors who who evaluate the election period for signs of each other's relative likelihood of winning a power struggle. Researchers who model election protest this way generally treat the use of election manipulation as a sign of the incumbent's unpopularity, and therefore weakness. Protest occurs after fraudulent elections in response to this perceived weakness; this possibility can serve to deter incumbents from engaging in manipulation under some circumstances. This is a normatively optimistic theory.

A second worldview argues that election manipulation can indicate a well-resourced, well-organized, powerful incumbent (independent of its popularity). Under this view, extensive election manipulation can be a show of strength, which induces cooperation with the incumbent from other societal actors. Since this view holds that more manipulation can further entrench the incumbent, it is a normatively pessimistic view.

Finally, others have argued that electoral protest occurs unrelated to strategic concerns about the relative strength of the incumbent and the opposition. In these models, electoral protest occurs because manipulation galvanizes citizens who hold diverse grievances, because economic conditions are poor, or simply because holding an election creates a moment of opportunity. The central feature of these models is that protest decisions are generally made without consideration for the likelihood that protests will overthrow the incumbent. As will be discussed in depth below, these different assumptions yield different predictions for when electoral protest is more likely to be initiated, and how long it will last against an incumbent that refuses to give up power. Table 1 presents these predictions for ease of reference.

```{=latex}
% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
% \usepackage{multirow}
\begin{table}[]
\begin{tabular}{@{}llll@{}}
\toprule
                                                                                                              & Manipulation signals: & \multicolumn{1}{c}{Strategic protest}                                                           & \multicolumn{1}{c}{Non-strategic protest}                                                  \\ \midrule
\multicolumn{1}{l|}{\multirow{2}{*}{Initiation most likely}}                                                  & Weakness              & \multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}Low margin + \\ high manipulation\end{tabular}}  & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Low margin + \\ high manipulation\end{tabular}} \\ \cmidrule(lr){2-3}
\multicolumn{1}{l|}{}                                                                                         & Strength              & \multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}Low margin + \\ low manipulation\end{tabular}}   &                                                                                            \\ \cmidrule(l){2-4} 
\multicolumn{1}{l|}{\multirow{2}{*}{Initiation least likely}}                                                 & Weakness              & \multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}High margin + \\ low manipulation\end{tabular}}  & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}High margin +\\ low manipulation\end{tabular}}  \\ \cmidrule(lr){2-3}
\multicolumn{1}{l|}{}                                                                                         & Strength              & \multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}High margin + \\ high manipulation\end{tabular}} &                                                                                            \\ \cmidrule(l){2-4} 
\multicolumn{1}{l|}{\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Duration when \\ incumbent wins\end{tabular}}} & Weakness              & \multicolumn{2}{c}{More manipulation: Longer protest}                                                                                                                                        \\
\multicolumn{1}{l|}{}                                                                                         & Strength              & \multicolumn{2}{c}{More manipulation: Shorter protest}                                                                                                                                       \\ \cmidrule(l){2-4} 
\end{tabular}
\caption{Predictions for protest initiation and duration under varying theoretical assumptions}
\label{tab:predictions}
\end{table}
```

@@

Fundamentally, the set of models which assume that protest is a centralized, strategic decision by opposition groups predict that the highest risk of protest should obtain when incumbents use massive fraud to generate a low incumbent vote-share, indicating that the incumbent's true popularity is very low. Conversely, they predict that protest should be least likely when the incumbent's vote-share is high and manipulation low, indicating a high level of true popularity that deters opposition protest.

The models described above conceive of opposition protest to election manipulation as a strategic action undertaken because the use of more manipulation increases the likelihood that the opposition will have support in the streets (due to the incumbent's low popularity). These are normatively optimistic models upon which a sizable literature has been founded: fraud can be kept in check, and ultimately reduced, through the development institutions that reveal information about election integrity: election monitoring [@Hyde2014;@daxecker2012], more independent courts [@chernykh2015third], media activity [@naiFourthEstate], and civil society coordination [@birch2017getting].

A second set of models shares the assumption that protest is a strategic action motivated by how confident opposition groups are they will succeed in a conflict with the incumbent. It differs, however, in its conceptualization of the information revealed by election manipulation. Most persuasively articulated by Simpser [-@simpser2013], this view holds that election manipulation demonstrates incumbent strength. As he writes, an incumbent that demonstrates the ability to engage in widespread manipulation "is also likely to
have the resources, capacities, and inclinations to overcome or punish opponents, reward supporters, and circumvent the law" [@simpser2013, p. 6]. A strong incumbent (in the language used by the first set of models, i.e. one that is relatively popular) may nonetheless manipulate heavily in order to produce a very large majority [@magaloni2006], to induce regime insiders to remain cooperative [@gehlbach2015electoral], or because of a bandwagon effect among front-line agents [@rundlett2016deliver]. By demonstrating otherwise difficult-to-observe capabilities, large-scale election manipulation can solve a coordination problem for the incumbent, securing the cooperation of elites [@simpser2013]. This can, in turn, fix expectations that the incumbent will continue to rule in the future, further enhancing its staying power [@hale_patronal_2014]. This argument is in keeping with Seeberg's [-@Seeberg2019] finding that higher-capacity states are associated with more election fraud.

Since these views hold that large-scale election manipulation can reveal information about sources of incumbent strength other than true popularity, and can have indirect, positive effects on incumbent stability, they are normatively pessimistic. They predict that opposition protest will be most likely when incumbent vote-share is low and manipulation is low, indicating that the incumbent lacks both genuine popularity and the resources and capacities needed to manufacture votes. By contrast, protest should be least likely to occur, under this framework, when incumbent vote-share and manipulation are both high.

Finally, a third school of thought holds that decisions to engage in electoral protest are non-centralized, driven by citizens' grievances rather than by a strategic, organized opposition [@Brancati2016]. In this view, election manipulation can act as a shared grievance among individuals who prefer to see the ruling party lose\textemdash especially when it appears that fraud may have been decisive [@kuntz_more_2009]. Belief in a stolen election is a particularly motivating kind of grievance, because it acts to focus all the diverse personal grievances individuals may have against the regime into one moment of opportunity for action [@tucker2007], possibly initiating a protest cascade [@Kuran1991]. These models thus tend to predict that protests are most likely when margins of victory are low and election manipulation high, suggesting both a high level of grievance in society and a "stolen" election. This framework predicts protest is least likely when incumbent vote-share is high and manipulation low (i.e. low grievances and low likelihood of a stolen election).

@@

@@

Since the empirical record is mixed, and two theories make similar predictions with regard to protest initiaiton, it is necessary to generate new hypotheses that can distinguish between the three models of election manipulation and protest that have been advanced. In particular, do post-election protest movements encounter weak incumbents as signaling models predict? If not, the deterrence model is undermined.

@@

Integrating Simpser's argument with recent research on the principal-agent dynamics inherent to election manipulation provides insights into this question. Where Simpser argues that manipulation should be blatant and excessive (that is, producing a large margin of victory) to send signals of strength, the principal-agent framework for understanding manipulation suggests otherwise. This research trend emphasizes the degree to which political principals require the cooperation of large numbers of front-line agents in order to generate extensive election fraud, agents who may shirk if their perceived interests are not aligned with the incumbent [@rundlett2016deliver]. Factors that minimize that principal-agent problem have been shown to facilitate greater election manipulation. These same factors\textemdash the incumbent's local popularity [@rundlett2016deliver], the incumbent's consolidated control over patronage resources [@Harvey2019@;@hale_patronal_2014], and low risk of punishment for pro-regime lawbreaking [@Harvey2022]\textemdash are also independently useful for suppressing post-election protest. Consequently, in an incumbent is able to manufacture a large numbers of votes, this is a signal of resource capacity, control over bureaucratic institutions, and perhaps deep wells of popularity in parts of the country. These signals remain independent sources of regime durability, even if the incumbent's margin of victory is low.

To the extent that election results in non-democracies send a signal, it is composed of a mix of three factors [@Pop-Eleches2015]: the incumbent's genuine popularity, its ability to mobilize otherwise reluctant voters (through patronage, clientelism, and coercion), and its ability to control the bureaucratic apparatus (including the criminal and judicial systems) to generate fraudulent votes. Large-scale election manipulation is enabled by consolidated control over patronage resources in society; this makes cooperation with the incumbent attractive for front-line agents, who are less willing to risk losing access to the incumbent's network [@Harvey2019]. However, such efforts also require that incumbents be able to shield front-line agents from negative consequences if their illegal efforts are publicly exposed [@Harvey2022]. In other words, the incumbent must have sufficient control over the police, the courts, and the media, such that ordinary individuals expect that punishment for illegal election manipulation is unlikely. At the same time, these capacities are useful for disarming protest movements, as discussed below.

The initial emergence of protest triggers a response by the government and ruling party that can take the form of repression, concessions, or toleration [@tillyMobilizationRevolution1978;@lipskyProtest1968;@Hummel2019;@raslerConcessionsRepression1996;@Bishara2015;@Frantz2014]. Concessions may be aimed at protesting parties themselves [@reuter2015;@mooreRepressionDissentSubstitution2000;@Hafner-Burton2016;@Piven1979] or the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019]. In either case, large-scale election manipulation indicates control over considerable patronage resources [@@Harvey2019@;@hale_patronal_2014;@greene_why_2007], which in turn indicates the capacity to engage in such payments (or withhold them from clients in exchange for passive support [@Rosenfeld2017]).

@@

In addition, as Gehlbach and Simpser [-@gehlbach2015electoral] argue, effective electoral manipulation can induce regime agents to exert effort on behalf of the incumbent. Elite defections, which can presage real challenges for the incumbent during a crisis, are also rarer in less democratic regions, where election manipulation is likely more severe [@Reuter2019]. Just as election manipulation has been held to solve a coordination problem for regime opponents, it may also solve a coordination problem for regime agents: when successful, a campaign to manufacture votes signals to agents that other agents are likely to stick with the incumbent due to its revealed capacities.

This argument\textemdash that successful, large-scale election manipulation draws on the same resources a state may use to suppress protest\textemdash turns the deterrent logic on its head. Instead of signalling weakness, large-scale fraud indicates a well of resilience: control over resources, impunity for pro-regime lawbreaking, and loyal agents. As a result, election outcomes contain information about several possible resources the incumbent may draw upon to defuse protests.

Prior models of electoral protest by strategic opposition actors treat election manipulation and genuine votes as substitute goods. For any given vote-share, an increase in manufactured votes reduces the number of genuine votes. For classic models, which consider manufactured votes to be merely illusions of strength, the higher the number of manufactured votes for any given vote total, the weaker the incumbent. A simplified version of this argument is presented in Figure \@ref(fig:classic). In this figure and subsequent one, dashed objects indicate variables that actors must estimate, solid items are observed, and ellipses indicate strategic actions. As the figure shows, opposition actors must estimate the proportion of manipulated versus genuine votes in the election outcome. The incumbent's genuine popularity drives both in the classic model, increasing genuine votes and reducing the need for manipulation. Since genuine popularity reduces the likelihood of a positive protest outcome for the opposition, genuine votes are taken as signals of incumbent strength. Manipulated votes are empty in the classic model; subtracting them from the incumbent's vote-total gives opposition an actors their estimate of the incumbent's ability to win a post-election standoff.

Thus the predicted relationship between manipulation and protest likelihood in the classic model is negative: more manipulation makes protest more likely. It is possible that that the size of the relationship is conditional on the incumbent's vote-share, however. If the incumbent's total vote-share is 70 percent, of which the opposition estimates 15 percent to be manufactured, the incumbent's genuine popularity is estimated at a respectable 55 percent. By contrast, if the incumbent posts a 50 percent margin of victory with the same 15 percent estimated to be manufactured, the opposition may perceive that the incumbent is weak in its genuine support (35 percent). This produces a first, baseline hypothesis, derived from the classic model.

> H1 (Strategic protest when manipulation signals weakness OR non-strategic protest): Election manipulation is positively associated with protest initiation, with the size of the effect increasing as incumbent vote-share declines.

It is important to note that the same hypothesis is also generated by non-centralized, non-strategic models of election protest. The mechanism, however, is different. In non-strategic models, protest emerges organically among disaffected citizens. Here, the combination of low incumbent vote-share and high manipulation is indicative of high grievances and a 'stolen' election. These are conditions that should be most likely to produce protest [@tucker2007].

```{r classic, echo=FALSE, fig.cap="A classic theory of election manipulation and protest", out.width="90%", fig.align='center'}
knitr::include_graphics("classic_theory.drawio.png")

@@

Alternatively, if strategic opposition groups recognize that large-scale election manipulation conveys information about regime durability, they should be deterred from initiating protest as the level of manipulation increases. This effect should be magnified by increasing vote-share, which suggests greater genuine popularity at any given level of manipulation.

> H2 (Strategic protest when manipulation signals strength): Election manipulation is negatively associated with protest initiation, with the size of the effect increasing as incumbent vote-share increases

Similarly, since the classic model holds that manipulated votes communicate little about incumbents' post-election strength, it implies that when incumbents rely heavily on election manipulation they will struggle to put down post-election protests. This effect should likewise be conditional on incumbent vote-share, since those incumbents who produce small margins of victory while relying heavily on manipulation will have limited ability to push back successfully against protest movements. Consequently, more election manipulation should be associated with longer-lived protest movements against surviving incumbents.^[In the appendix, I check the separate prediction that manipulation should be positively associated with incumbents being ousted by protest movements.]

> H3 (manipulation signals weakness): Election manipulation is positively associated with protest movement duration, with the size of this effect decreasing with incumbent vote-share.

```{r newtheory, echo=FALSE, fig.cap="Theory diagram", out.width="90%", fig.align='center'}
knitr::include_graphics("theory_diagram.drawio.png")

Lastly, if election manipulation is an indicator of underlying capacities the incumbent may use to defeat protest movements, this theory predicts that more severe electoral manipulation will be associated with shorter-lived protest waves. Incumbents with the resources and institutions needed to produce widespread election manipulation can draw on those same capacities to buy off, repress, or otherwise quash post-election protest.

> H4 (manipulation signals strength): Among incumbents that survive post-election protest movements, more extensive election fraud should be associated with shorter protest waves.

Figure \@ref(fig:newtheory) illustrates how the present argument differs from the classic model. Manufactured votes are not negatively associated with genuine popularity, but positively associated with impunity for agents and control over patronage resources. Abundant resources and extensive impunity reduce the principal-agent problems that frustrate election manipulation efforts. These same resources, institutions, and loyal agents make protest outcomes less likely to be favorable for the opposition. As such, opposition actors interpret manufactured and genuine votes as information about separate channels of incumbent strength. This produces hypotheses opposite to those of the classic model. Manipulation should be associated with reduced likelihood of protest, since it is a signal of capabilities that make protest success less likely. This effect should increase as margins of victory increase for the incumbent, providing additional, distinct information about incumbent resilience.

The argument can be summarized as follows. Incumbents will generally seek to manipulate elections as much as they can, to capitalize on both the direct and indirect benefits of electoral manipulation [@simpser2013]. To do so, they must solve principal-agent problems by consolidating control over patronage resources and shielding agents from negative consequences of their actions. The election outcome sends an ambiguous signal of the incumbents' resources\textemdash including natural popularity, resource control, and control over the legal apparatus\textemdash to possible regime opponents. The more successfully the incumbent has solved the principal-agent problems associated with election manipulation, however, the more likely it will be able to use those tools to quash protest. As a result, more extensive manipulation should be associated with shorter-lived protest movements against incumbents.

@@

ECAV records protests and other events related to an election, ranging from six months before to three months after the vote (or three months after the final round of a multi-round election). To qualify, an action must be contentious; rallies in support of the incumbent are excluded. I only include events labeled as protests, occupations, or blockades, excluding acts like shootings, bombings, and coups. ECAV data is in event-day format; to capture the *number of protests* before the election, I take the sum number of events recorded in ECAV before election day. To evaluate hypotheses regarding *protest duration* I take the number of days between the last protest recorded in ECAV and the first, for each election.^[The duration of a protest wave with only one protest is taken to be one day. While the main analysis assumes that the last protest in the dataset marks the end of the protest movement, an alternative specification in the appendix assumes right-censoring for a portion of the observations due to the 90-day observation window.]

>>>>

The stakes of this debate are significant. The former view is an optimistic one; incumbents avoid fraud when it might be exposed and reveal their weakness to strategic protesters\textemdash an auspicious prediction for democratization and ultimately the under-pinning of 'self-enforcing' democracy [@fearon_self-enforcing_2011]. If the latter view holds, electoral protest may mainly occur non-strategically, and would typically fail to deter election manipulation as a result. Mass protest would, in this case, be a fragile guardrail against autocratization. This debate cannot be advanced by focusing solely on protest occurrence. Instead, this paper contributes by drawing out implications from three strands of the literature on election manipulation and protest, which have (to my knowledge) not previously been explicated.

In short: there are two views on the nature of electoral protest which make the same prediction. If opposition groups strategically initiate protest when they believe the incumbent to be weak, they will be most likely to do so when the election results show a low vote-share for the incumbent combined with high levels of manipulation\textemdash a combination of factors which reveals the incumbent's very low genuine popularity. [NOTE: ADD CITES HERE] However, if protest movements are more organic and non-strategic, reflecting citizen grievances, we should expect protest initiation in the same circumstances. Here the combination of low incumbent vote-share and high manipulation indicates severe citizen grievances (low natural popularity) which are galvanized by the shared grievance of a 'stolen' election [@tucker2007]. Since both theories make the same prediction about protest initiation, such tests\textemdash which have been the dominant empirical measure in the literature\textemdash do not reveal whether protest is strategic and manipulation indicates weakness or if manipulation is a hallmark of strong incumbents [@simpser2013] and protest is non-strategic.

To unravel this question, I take this analysis one step further. In strategic-protest models, the opposition protests because it believes the probability of a pro-government outcome of the protest is low. I assume that the duration of protest movements, when they begin, is an indicator of this underlying likelihood of a pro-government outcome. The more quickly a government can put down protests, the greater the underlying (unobserved) distribution of power favors the incumbent over the opposition. By contrast, the longer it takes the government to restore control, the less favorable the distribution of power.^[The analysis treats the small number of cases where the opposition succeeds in ousting the government in several ways. They are excluded from the main analysis, but incorporated in multiple robustness checks.]

I make use of data on the duration of post-election protests. Shorter-lived protest movements that end with the incumbent still in power indicate a government that is relatively strong compared to the protesting opposition, while longer-lived protest movements indicate a more even balance of power between the two. The reverse is also true: shorter-lived protest movements that end in opposition victory indicate a weaker government, while governments that hold on longer before being defeated are relatively stronger. Using data from 614 non-democratic elections across 97 countries, from 1989 to 2012, this paper finds that 1) protest initiation is most likely when manipulation is severe but incumbent vote-share is low, as both theories predict, and 2) that governments that engage in more manipulation are likely to put down protests more quickly than governments that produce little manipulation, especially when incumbents claim small vote-shares.

@@

@@

Fundamentally, both set of models share the assumption that protest is a centralized, strategic decision by opposition groups, and that fraud is a tool employed by weak (i.e. unpopular) incumbents. As a result, they predict that the highest risk of protest should obtain when incumbents use massive fraud but still produce a low incumbent vote-share, indicating that the incumbent's true popularity is very low. This, in turn, indicates to opposition actors that protest is more likely to succeed in unseating the incumbent. These are normatively optimistic models upon which a sizable literature has been founded: fraud can be kept in check, and ultimately reduced, through the development of institutions that reveal information about election integrity: election monitoring [@Hyde2014;@daxecker2012], more independent courts [@chernykh2015third], media activity [@naiFourthEstate], and civil society coordination [@birch2017getting]. These institutions improve election integrity, it is claimed, by increasing the likelihood that fraud will be observed and thus, incumbent weakness exposed.

However, an alternative set of models have also been advanced. A second worldview argues that election manipulation can indicate a well-resourced, well-organized, powerful incumbent (independent of its popularity). Under this view, extensive election manipulation can be a show of strength, which induces cooperation with the incumbent from other societal actors. Since this view holds that more manipulation can further entrench the incumbent, it is a normatively pessimistic view.

Most persuasively articulated by Simpser [-@simpser2013], this view holds that election manipulation demonstrates incumbent strength. As he writes, an incumbent that demonstrates the ability to engage in widespread manipulation "is also likely to have the resources, capacities, and inclinations to overcome or punish opponents, reward supporters, and circumvent the law" [@simpser2013, p. 6]. A strong incumbent (in the language used by the first set of models, i.e. one that is relatively popular) may nonetheless manipulate heavily in order to produce a very large majority [@magaloni2006], to induce regime insiders to remain cooperative [@gehlbach2015electoral], or because of a bandwagon effect among front-line agents [@rundlett2016deliver]. By demonstrating otherwise difficult-to-observe capabilities, large-scale election manipulation can solve a coordination problem for the incumbent, securing the cooperation of elites [@simpser2013]. This can, in turn, fix expectations that the incumbent will continue to rule in the future, further enhancing its staying power [@hale_patronal_2014]. This argument is in keeping with Seeberg's [-@Seeberg2019] finding that higher-capacity states are associated with more election fraud.

For the set of theories that treat election manipulation as an indicator of sources of regime strength, manipulation can serve to deter protest under some circumstances. [NOTE ADD CITES HERE?] This leaves protest, when it does occur in the presence of large-scale manipulation, as something of a puzzle if protest behavior is largely strategic. For this reason, the manipulation-as-strength view comports more easily with another established view of the causes of electoral protest. Researchers in this tradition have argued that electoral protest occurs unrelated to strategic concerns about the relative strength of the incumbent and the opposition. In these models, electoral protest occurs because manipulation galvanizes citizens who hold diverse grievances, because economic conditions are poor, or simply because the bare fact of holding an election creates a moment of opportunity for opposition organization irrespective of the level of manipulation [@Pop-Eleches2015] [NOTE: ADD CITES]. The central feature of these models is that protest decisions are generally made without consideration for the likelihood that protests will overthrow the incumbent. As will be discussed in depth below, these different assumptions yield different predictions for when electoral protest is more likely to be initiated, and how long it will last against an incumbent that refuses to give up power.

This school of thought holds that decisions to engage in electoral protest are non-centralized, driven by citizens' grievances rather than by a strategic, organized opposition [@Brancati2016]. In this view, election manipulation can act as a shared grievance among individuals who prefer to see the ruling party lose\textemdash especially when it appears that fraud may have been decisive [@kuntz_more_2009]. Belief in a stolen election\textemdash one the incumbent would have lost in the absence of manipulation\textemdash is a particularly motivating kind of grievance, because it acts to focus all the diverse personal grievances individuals may have against the regime into one moment of opportunity for action [@tucker2007], possibly initiating a protest cascade [@Kuran1991]. These models thus tend to predict that protests are most likely when margins of victory are low and election manipulation high, suggesting both a high level of grievance in society and a "stolen" election. This framework predicts protest is most likely when incumbent vote-share is low and manipulation high (i.e. elevated grievances and a high likelihood of a stolen election).

@@

```{=latex}
% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\begin{tabular}{@{}lllllll@{}}
\toprule
                     &              &                           & \multicolumn{2}{l}{Strategic protest; manipulation = weakness} & \multicolumn{2}{l}{Non-strategic protest; manipulation = strength} \\ \midrule
Incumbent vote-share & Manipulation & True popularity inference & Initiation              & Prob. of pro-gov outcome             & Initiation                & Prob. of pro-gov. outcome              \\
Low                  & Low          & Low                       & Less likely             & More likely                          & Less likely               & Less likely                            \\
Low                  & High         & Very low                  & More likely             & Less likely                          & More likely               & More likely                            \\
Citations            &              &                           &                         &                                      &                           &                                        \\ \bottomrule
\end{tabular}
\caption{}
\label{tab:predictions}
\end{table}
```

@@

Since the empirical record is mixed, and two theories make similar predictions with regard to protest initiation, it is necessary to generate new hypotheses that can distinguish between the two models of election manipulation and protest that have been advanced. In particular, do post-election protest movements encounter weak incumbents as signaling models predict? If not, the deterrence model is undermined.

@@

Integrating Simpser's argument with recent research on the principal-agent dynamics inherent to election manipulation provides insights into this question. Where Simpser argues that manipulation should be blatant and excessive (that is, producing a large margin of victory) to send signals of strength, the principal-agent framework for understanding manipulation suggests otherwise. This research trend emphasizes the degree to which political principals require the cooperation of large numbers of front-line agents in order to generate extensive election fraud, agents who may shirk if their perceived interests are not aligned with the incumbent [@rundlett2016deliver]. Factors that minimize principal-agent problems have been shown to facilitate greater election manipulation. These include the incumbent's local popularity [@rundlett2016deliver], the incumbent's ability to distribute patronage resources [@Harvey2019;@hale_patronal_2014], and low risk of punishment for pro-regime lawbreaking [@Harvey2022]. As discussed below, these are also independently useful for suppressing post-election protest. Consequently, if an incumbent is able to manufacture a large numbers of votes, it is also likely to have the resource capacity, control over bureaucratic institutions, and perhaps deep wells of popularity in parts of the country that can serve as independent sources of regime durability, even if the incumbent's nationwide margin of victory is low.

The initial emergence of protest triggers a response by the government and ruling party that can take the form of repression, concessions, or toleration [@tillyMobilizationRevolution1978;@lipskyProtest1968;@Hummel2019;@raslerConcessionsRepression1996;@Bishara2015;@Frantz2014]. Concessions may be aimed at protesting parties themselves [@reuter2015;@mooreRepressionDissentSubstitution2000;@Hafner-Burton2016;@Piven1979] or the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019]. In either case, large-scale election manipulation indicates control over considerable patronage resources [@Harvey2019;@hale_patronal_2014;@greene_why_2007], which in turn indicates the capacity to engage in such payments (or withhold them from clients in exchange for passive support [@Rosenfeld2017]).

@@

In addition, as Gehlbach and Simpser [-@gehlbach2015electoral] argue, effective electoral manipulation can induce regime agents to exert effort on behalf of the incumbent. Elite defections, which can presage real challenges for the incumbent during a crisis, are also rarer in less democratic regions, where election manipulation is likely more severe [@Reuter2019]. Just as election manipulation has been held to solve a coordination problem for regime opponents, it may also solve a coordination problem for regime agents: when successful, a campaign to manufacture votes signals to agents that other agents are likely to stick with the incumbent due to its revealed capacities [@hale_patronal_2014].

This argument\textemdash that successful, large-scale election manipulation draws on the same resources a state may use to suppress protest\textemdash turns the deterrent logic on its head. Instead of signalling weakness, large-scale fraud indicates a well of resilience: control over resources, impunity for pro-regime lawbreaking, and loyal agents. As a result, governments that build and deploy an electoral manipulation apparatus are simultaneously innoculating themselves against the cost of protest.

Prior models of electoral protest by strategic opposition actors treat election manipulation and genuine votes as substitute goods. For any given vote-share, an increase in manufactured votes reduces the number of genuine votes. For classic models, which consider manufactured votes to be merely illusions of strength, the higher the number of manufactured votes for any given vote total, the weaker the incumbent. This means that strategic opposition actors are more likely to initiate protest when they perceive the incumbent to be weaker, and expect to be more likely to succeed in a challenge. A simplified version of this argument is presented in Panel B of Figure \@ref(fig:theoryfig). In the figure, dashed objects indicate variables that actors must estimate, while solid items are observed.

As the figure shows, in the classic model, the incumbent's (unknown) true popularity is positively related to its vote-share in the election, and negatively related to its use of fraud. This dynamic is the fulcrum upon which these models hinge: less popular incumbents (who are expected to be less likely to defeat an opposition protest) must manipulate more in order to win elections. Popular incumbents need manipulate less. Opposition actors gauge the probability of an pro-government protest outcome by estimating the incumbent's true popularity based on the degree of manipulation and the incumbent's vote-share in the election. The level of manipulation is, in effect, a mediator for the effect of vote-share. A high vote-share with low fraud indicates a more popular incumbent, likely deterring protest. A low vote-share with a high degree of fraud indicates a very weak incumbent, making strategic protest more likely. In other words, manipulated votes are empty in the classic model; subtracting them from the incumbent's vote-total gives opposition actors their estimate of the incumbent's ability to win a post-election standoff.

Thus the predicted relationship between manipulation and protest likelihood in the classic model is positive: more manipulation makes protest more likely. However, as increasing vote-share increases the number of genuine votes for the incumbent at any fixed level of fraud, the effect of fraud should be conditional on the claimed vote-share [NOTE: ADD CITES]. If the incumbent's total vote-share is 70 percent, for example, of which the opposition estimates 15 percent to be manufactured, the incumbent's genuine popularity is estimated at a respectable 55 percent. By contrast, if the incumbent posts a 50 percent margin of victory with the same 15 percent estimated to be manufactured, the opposition may perceive that the incumbent is weak in its genuine support (35 percent). This produces a first, baseline hypothesis, derived from the classic model.

> H1: Increasing election fraud makes protest more likely, with the size of this effect decreasing with incumbent vote-share.

It is important to note that the same hypothesis is also generated by non-centralized, non-strategic models of election protest. The mechanism, however, is different. In non-strategic models, protest emerges organically among disaffected citizens. Here, the combination of low incumbent vote-share and high manipulation is indicative of high grievances and a 'stolen' election. These are conditions that should be most likely to produce protest [@tucker2007]. By contrast, as vote-share increases for a set level of fraud, the incumbent's genuine popularity is higher and thus grievances are lower (reducing the likelihood of protest).

```{r theoryfig, echo=FALSE, fig.cap="A classic theory of election manipulation and protest", out.width="90%", fig.align='center'}
knitr::include_graphics("theory_comp.drawio.png")

@@

The classic model holds that manipulated votes correlate negatively with incumbents' post-election strength, implying that when incumbents rely heavily on election manipulation they will struggle to put down post-election protests. This effect should likewise be conditional on incumbent vote-share, since those incumbents who produce small margins of victory while relying heavily on manipulation will have limited ability to push back successfully against protest movements. Panel B in Figure [NOTE: FIX REF] illustrates this; a higher likelihood of pro-government protest outcome reduces the likelihood that the opposition will initiate protest. Consequently, classic models predict that more election manipulation should be associated with longer-lived protest movements against surviving incumbents.^[In the appendix, I check the separate prediction that manipulation should be positively associated with incumbents being ousted by protest movements.]

On the other hand, if election manipulation is an indicator of underlying capacities the incumbent may use to defeat protest movements, more severe electoral manipulation will be associated with shorter-lived protest waves. Panel A of Figure \@ref(fig:theoryfig) illustrates how the present argument differs from the classic model. Manufactured votes are not negatively associated with genuine popularity, but positively associated with impunity for agents and control over patronage resources. They are also positively associated with incumbent popularity [@rundlett2016deliver]. Abundant resources and extensive impunity reduce the principal-agent problems that frustrate election manipulation efforts. These same resources, institutions, and loyal agents make protest outcomes less likely to be favorable for the opposition. Note that, in this model, protest initiation is not tied to expected protest outcome; protest is driven by non-strategic concerns related to the incumbent's popularity (i.e. citizens' diverse grievances) and whether manipulation is perceived as having been decisive (i.e. a stolen election). Incumbents with the resources and institutions needed to produce widespread election manipulation can draw on those same capacities to buy off, repress, or otherwise quash post-election protest. This yields Hypothesis 2.

> H2: Election manipulation is positively associated with protest movement duration, with the size of this effect decreasing with incumbent vote-share.

The argument can be summarized as follows. Incumbents will generally seek to manipulate elections as much as they can, to capitalize on both the direct and indirect benefits of electoral manipulation [@simpser2013]. To do so, they must solve principal-agent problems by consolidating control over patronage resources and shielding agents from negative consequences of their actions. The same resources that can be used to manipulate elections on a wide scale\textemdash including natural popularity, resource control, and control over the legal apparatus\textemdash also better enable incumbents to dismantle protest movements when they do emerge. As a result, more extensive manipulation should be associated with shorter-lived protest movements against incumbents. Much election-related protest, then, is unrelated to strategic concerns about the likelihood of protest success.

@@

ECAV records protests and other events related to an election, ranging from six months before to three months after the vote (or three months after the final round of a multi-round election). To qualify, an action must be contentious; rallies in support of the incumbent are excluded. I only include events labeled as protests, occupations, or blockades, excluding acts like shootings, bombings, and coups. ECAV data is in event-day format; to capture the *number of protests* before the election, I take the sum number of events recorded in ECAV before election day. To evaluate hypotheses regarding *protest duration* I take the number of days between the last protest recorded in ECAV and the first, for each election.^[The duration of a protest wave with only one protest is taken to be one day.]

The large majority of elections experience no post-election protest: in `r nelda2.sub %>% filter(n.events.post == 0) %>% nrow()` of the `r nrow(nelda2.sub)` elections, no protests are recorded. Figure \@ref(fig:elex-hist) reports the distribution of election fraud across the dataset, as well as protest duration. As the figure shows, while election fraud is common among this set of regimes, protests are rare, and, when they do occur, are usually short in duration. This distribution\textemdash with large numbers of zeroes\textemdash is in line with the theoretical proposition that most fraudulent elections do not produce signals of weakness that prompt opposition protest. Still, approximately one-third of elections experienced some protest, and so protest initiation is modeled using standard logistic regression. Protest wave duration is modeled using Cox proportional hazards models.
